Gradual Soundness: Lessons from Static Python

### 2021-12-07 pj-7.1/paper.tex

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle soundly (FILL examples ... while loop?);
omitting these lets us focus on the difficult corners.
Other Python features are left out because \SP{} does not ascribe types
to them.
These include \code{eval}, first-class classes, and module-level cells;
see \cref{s:impl} for further discussion.

@@

%% SP side channels:
%% eval,
%% the locals and globals dicts (the latter can be accessed through frame objects at the moment),
%% and mutable closure cells.

@@

\SP{} does not attempt to type first-class classes:
partly because they see little use at Instagram (FILL how much?)
and partly because it is unclear how to incorporate them into
the nominal type system.
The straightforward but restrictive approach is to force code
that uses a first-class class to expect instances of a particular
static class.
FILL is the bracha cook mixin pattern a way to work within nominal~\cite{bc-oopsla-1990}?
\cite{fkf-popl-1998} propose a different method that accommodates mixins,
but the method calls a second layer of interface types.
\cite{mt-oopsla-2021} similarly allow structural objects with interfaces.
Adding these layers to \SP{} is an open question; in particular,
prior work does not handle multiple inheritance.

@@

>>>>

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle soundly (FILL examples ... while loop?);
omitting these lets us focus on the difficult corners.
Other Python features are left out because their \SP{} semantics is identical
to Python.
These include \code{eval}, first-class classes, multiple inheritance, and module-level cells.
\SP{} does not rely on types for these features to guide program transformations;
see \cref{s:impl} for further discussion.

@@

%% TODO discuss optimizations here, not in sec 2

%% TODO SP side channels:
%% ?? eval,
%% the locals and globals dicts (the latter can be accessed through frame objects at the moment),
%% and mutable closure cells.

@@

\SP{} does not attempt to type first-class classes:
partly because they see little use at Instagram (FILL how much?)
and partly because it is unclear how to incorporate them into
the nominal type system.
The straightforward but restrictive approach is to force code
that uses a first-class class to expect instances of a particular
static class.
FILL is the bracha cook mixin pattern a way to work within nominal~\cite{bc-oopsla-1990}?
Flatt et al.~\cite{fkf-popl-1998} propose a different method that accommodates mixins,
but the method calls a second layer of interface types.
Muehlboeck and Tate~\cite{mt-oopsla-2021} similarly allow structural objects with interfaces.
Adding these layers to \SP{} is an open question; in particular,
prior work does not handle multiple inheritance.

@@

\paragraph{Multiple Inheritance}

%% https://docs.python.org/3/reference/datamodel.html

Python allows multiple inheritance.
Classes can inherit from any number of other classes.
To resolve method calls, Python traverses the class hierarchy
to find the first matching receiver (MRO).

[FILL]
\SP{} lets most programs use multiple inheritance.
The one constraint is that a class cannot have two
values bound to the same member name, to ensure v-table
integrity.
